Synthesis of Linear Functional Programs

### 2021-07-11 article/article.tex
Write on recursive ADTs

\newcommand{\synname}{\emph{SILI}}
\newcommand{\te}[1]{\textrm{\emph{#1}}}

@@

@@

    % -o R
    \infer*[right=($\lolli R$)]
    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \vdash M : B \Uparrow \and x
    \not\in \Delta'}
    {\Gamma ; \Delta/\Delta' ; \Omega \vdash \lambda x \rightarrow M : A
    \lolli B \Uparrow}

@@

\begin{mathpar}
    \infer*[right=($\Uparrow R$)]
    {\Gamma ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash M : C \and C\ \textrm{not
    right asynchronous}}
    {\Gamma ; \Delta/\Delta' ; \Omega \vdash M : C \Uparrow}
\end{mathpar}

@@

    \infer*[right=($\tensor L$)]
    {\Gamma ; \Delta/ \Delta' ; \Omega, y{:}A, z{:}B \Uparrow\ \vdash M : C
    \and y,z \not\in \Delta'}
    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \tensor B \Uparrow\ \vdash\
    \textrm{let}\ y \tensor z = x\ \textrm{in}\ M : C}

@@

    \mprset{flushleft}
    \infer*[right=($\oplus L$)]
    {
    \Gamma ; \Delta/ \Delta' ; \Omega, y{:}A \Uparrow\ \vdash M : C \and
    y \not\in \Delta' \\
    \Gamma ; \Delta/ \Delta'' ; \Omega, z{:}B \Uparrow\ \vdash N : C \\
    z \not\in \Delta'' \\
    \Delta' = \Delta''
    }
    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \oplus B \Uparrow\ \vdash\
    \textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y \rightarrow M\ |\
    \textrm{inr}\ z \rightarrow N : C}

@@

\begin{mathpar}
    \infer
    {\vdots}
    {\Gamma, \textrm{\emph{foo}}{:}A \lolli B; \Delta/\Delta'; \Omega \vdash M :
    A \lolli B \Uparrow}
\end{mathpar}
and, by definition, all subsequent inference rules will have
($\textrm{\emph{foo}}{:}A \lolli B$) in the $\Gamma$ context too.

We can force the usage of the recursive call by adding it not only to the
unrestricted context, but to the linear one as well.

@@

\begin{mathpar}
    \infer*[right=(adtR)]
    {\Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{Bar}_n \ M : \te{Foo} \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    { \Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C \and
    y_1 \not\in \Delta'_1
    \\ \Gamma ; \Delta/ \Delta'_2 ; \Omega, y_2{:}X_2 \Uparrow\ \vdash M_2 : C
    \\ y_2 \not\in \Delta'_2
    \\\\ \\ \dots
    \\ \Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
    \\ y_n \not\in \Delta'_n
    \\ \Delta'_1 = \Delta'_2 = \dots = \Delta'_n }
    {\Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}\te{Foo} \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ \textrm{Bar}_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}

where the ADT \emph{Foo} and its definition stand for any ADT with form
\[
    \textrm{data Foo} = \textrm{Bar}_1\ X_1\ \vert\ \textrm{Bar}_2\ X_2\ \vert\
\dots\ \vert\ \textrm{Bar}_n\ X_n
\]

@@

>>>>

\newcommand{\synname}{\emph{SILI}}
\def\Rho{P}
\newcommand{\te}[1]{\textrm{\emph{#1}}}

@@

This section presents the rules while explaining them, modifying some along the
process. For a

@@

    % -o R
    \infer*[right=($\lolli R$)]
    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \vdash M : B \Uparrow \and x
    \notin \Delta'}
    {\Gamma ; \Delta/\Delta' ; \Omega \vdash \lambda x \rightarrow M : A
    \lolli B \Uparrow}

@@

\begin{mathpar}
    \infer*[right=($\Uparrow R$)]
    {\Gamma ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash C \and C\ \textrm{not
    right asynchronous}}
    {\Gamma ; \Delta/\Delta' ; \Omega \vdash C \Uparrow}
\end{mathpar}

@@

    \infer*[right=($\tensor L$)]
    {\Gamma ; \Delta/ \Delta' ; \Omega, y{:}A, z{:}B \Uparrow\ \vdash M : C
    \and y,z \notin \Delta'}
    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \tensor B \Uparrow\ \vdash\
    \textrm{let}\ y \tensor z = x\ \textrm{in}\ M : C}

@@

    \mprset{flushleft}
    \infer*[right=($\oplus L$)]
    {
    \Gamma ; \Delta/ \Delta' ; \Omega, y{:}A \Uparrow\ \vdash M : C \and
    y \notin \Delta' \\
    \Gamma ; \Delta/ \Delta'' ; \Omega, z{:}B \Uparrow\ \vdash N : C \\
    z \notin \Delta'' \\
    \Delta' = \Delta''
    }
    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \oplus B \Uparrow\ \vdash\
    \textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y \rightarrow M\ |\
    \textrm{inr}\ z \rightarrow N : C}

@@

\begin{mathpar}
    \infer
    {\dots}
    {\Gamma, \textrm{\emph{foo}}{:}A \lolli B; \Delta/\Delta'; \Omega \vdash M :
    A \lolli B \Uparrow}
\end{mathpar}
and, by definition, all subsequent inference rules will have
($\textrm{\emph{foo}}{:}A \lolli B$) in the $\Gamma$ context too.

We can also force the usage of the recursive call by adding it not only to the
unrestricted context, but to the linear one as well.

@@

\begin{mathpar}
    \infer*[right=(adtR)]
    {\Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{Bar}_n \ M : \te{Foo} \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        \Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \and
        y_1 \notin \Delta'_1
        \\
        \Gamma ; \Delta/ \Delta'_2 ; \Omega, y_2{:}X_2 \Uparrow\ \vdash M_2 : C
        \\
        y_2 \notin \Delta'_2
        \\\\
        \\ \dots
        \\
        \Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \\
        y_n \notin \Delta'_n
        \\
        \Delta'_1 = \Delta'_2 = \dots = \Delta'_n
    }
    {\Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}\te{Foo} \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ \textrm{Bar}_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}

where the ADT \emph{Foo} and its constructors stand for any ADT with form
\[
    \textrm{data Foo} = \textrm{Bar}_1\ X_1\ \vert\ \textrm{Bar}_2\ X_2\ \vert\
\dots\ \vert\ \textrm{Bar}_n\ X_n
\]

@@

To handle this situation, we impede the decomposition of an ADT in subsequent
proofs of its branches, and to impede the construction of an ADT in a subsequent
proof of a type to construct it. For this, we need two more contexts, $\Rho_C$
for restrictions on construction and $\Rho_D$ for restrictions on
deconstruction. For convenience, they are represented by a single $\Rho$ if
unused. All rules must trivially propagate these. The ADT rules are explicitly
modified to take this into consideration:

\begin{mathpar}
    \infer*[right=(adtR)]
    {(\Rho_C'; \Rho_D) ; \Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow \and
    \te{Foo} \notin \Rho_C}
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta' \vdash\ \textrm{Bar}_n \ M : \te{Foo} \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        \te{Foo} \notin \Rho_D
        \and
        \Delta'_1 = \dots = \Delta'_n 
        \\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \and
        y_1 \notin \Delta'_1
        \\\\
        \\ \dots
        \\\\
        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \and
        y_n \notin \Delta'_n
    }
    {(\Rho_C; \Rho_D); \Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}\te{Foo} \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ \textrm{Bar}_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}
where
\begin{mathpar}
    \Rho'_C = \textrm{\textbf{if}}\ \te{Foo}\ \textrm{is recursive \textbf{then}}\ \Rho_C,
    \te{Foo}\ \textrm{\textbf{else}}\ \Rho_C

    \Rho'_D = \textrm{likewise}
\end{mathpar}

These modifications prevent the derivation from not ending for the two seen
situations, however, they also greatly limit the derivable valid programs space,
leaving the synthetiser pretty much unable to synth specifications with
recursive types, which is so undesirable that we add two rules to complement the
restrictions on construction and destruction of recursive types.

First, after decomposing a recursive ADT, we must consume all constructor
parameters, and since we can't decompose the self type any further because of a
restriction, we should be able to use it from the linear context; even though,
non-restricted types are not to take that course since they weren't unnaturally
restricted, and because it would result in additional non-determinism and
infinite recursion. Second, without any additional rules, an ADT in the linear
context will loop back to the inversion context, and then loop again, going back and
forth between the two contexts for infinity; instead, when focused on an ADT, we should
either instance the goal (provided they're the same type), fail if the
decomposition is restricted, or switch to inversion if it's not. The three
following rules ensure this:

\begin{mathpar}
    \infer*[right=(adt$\Uparrow$L)]
    {
        (\Rho_C; \Rho_D);\Gamma; \Delta, x{:}\te{Foo}/\Delta'; \Omega \Uparrow\ \vdash M : C
        \and
        \te{Foo} \in \Rho_D
    }
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \Uparrow\ \vdash M : C}
    \and
    \infer*[right=(adt-init)]
    {  }
    {\Rho; \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Downarrow\ \vdash x : \te{Foo}}
    \and
    \infer*[right=(adt$\Downarrow$L)]
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Uparrow\ \vdash M :
    \te{Foo} \and \te{Foo} \notin \Rho_D}
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Downarrow\ \vdash M : \te{Foo}}
\end{mathpar}

% TODO: Adicionar regra que diz que se for ADT Rec -o A então a construção de
% ADT Rec é logo restrita?? parece que acelera minimamente mas não arranjo um
% exemplo em que seja crucial. pode ser uma coisa que tenha ficado para resolver
% uma coisa antiga que ficou resolvida mais tarde com outra modificação e então
% é agora inutil. Talvez seja demais adicionar à regra

conclude...
